Data Specialization for a Staged Lambda Calculus

### 2014-11-11 paper/intro.tex
major reorganization of section 2 (in progress) - moving lots of text around!

>>>>

\crem{This text was lifted from section 2.}

Prior work on metaprogramming allows us to evaluate such staged code
in a staged fashion by first providing the first-stage arguments and
generating a piece of code that is specialized for the arguments.  For
example, we can evaluate the staged quickselect by providing a
list~\texttt{l} to obtain a piece of code that then can be used with
different ranks (\texttt{k} values) in the second stage.  Such a
specialized code essentially performs all the first-stage computations
first and uses their results to improve the efficiency in the second
stage, where we issue many queries.

This staged evaluation technique corresponds to {\em partial
  evaluation} or {\em program specialization} and can bring out
dramatic improvements in the efficiency of programs.  It has however
one major limitation: since it requires the first stage arguments to
be available, it cannot be performed at compile time, except of course
when the first-stage values are available at compile time.  Since the
evaluation in the first stage produces a piece of code that must also
be compiled or interpreted to execute in the second stage, the
approach can lead to code blowup and tends to be relatively expensive
in practice.  Due to these reasons, the applications of partial
evaluation are limited.  For example, in our quickselect example, it
is unlikely to know the list~\texttt{l} that we will be operation on
at compile time. Partial evaluation dynamically at run time is
difficult because the first stage generates code that is linear in the
size of the list, whose generation, dynamic compilation or
interpretation can be excruciatingly slow.

\cite{knoblock96} proposed the {\em data specialization}
technique to overcome some of these limitations of program
specialization.  The idea behind this technique is to take a staged
program and split it into a stage-1 and a stage-2 program without
knowledge of any of the arguments.  Unlike program specialization,
data specialization does not require knowledge of the first-stage
arguments to produce fast code for the second stage.  Instead it yield
a first stage program that maps the first stage arguments to an
intermediate or {\em boundary data structure}, which is used by the
second stage program along with the second stage arguments to produce
the final results.  Knoblock and Ruf's observation is that the
information required by the second stage can be represented solely as
a data structure, making it unnecessary to generate and compile or
interpret code dynamically at run time.  Data specialization can
therefore be applied statically at compile time, leading to
drastically improved run-time performance without any significant
run-time overheads.

It is thus perhaps no surprise that the prior art in data
specialization is not able to perform such ``optimizations''.
Knoblock and Ruf's paper takes a very important step by formulating
the problem and proposing a solution.  But their work applies to
program fragments that don't involve sophisticated data types such as
as sums and recursive types, and don't involve recursion. For example,
their work would not apply to our quickselect example and similar ones
involving recursion and/or  recursive types.

\subsection{Old intro}

### 2014-11-11 paper/semantics.tex
major reorganization of section 2 (in progress) - moving lots of text around!

>>>>

\crem{Fold in the next paragraph:}

We start by defining a simple staged language, called \lang, to express the
input to our splitting algorithm.  \lang\ unambiguously specifies how its terms
should be split
% and features a type system that ensures valid splittings exist.
% only minimal refactoring should be required to write terms within this language,
%and from there, the type system should be enough to prove that a valid splitting exists.
by identifying each with one of two {\em stages}, namely \bbone\ or \bbtwo.
Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate
it---all stage-\bbone\ subexpressions are evaluated before stage-\bbtwo\ ones.
After splitting, those parts of the \lang\ term that were identified with stage
\bbone\ will end up in the stage \bbone\ function ({\tt qSelect1}), and those
parts identified with stage \bbtwo\ will end up in the stage \bbtwo\ function
({\tt qSelect2}).  Typing rules for \lang\ ensure a valid splitting exists.
That is, within well-typed terms, information can flow from stage \bbone\ to
stage \bbtwo\ portions, but never from \bbtwo\ to \bbone.
